A coherent and comprehensive dating system for stars in the Milky Way

### 2017-05-03 paper/ms.tex
adding words drunk on the train

\begin{abstract}
% Aims
Stellar age underpins a wide range of astronomical applications, yet
remains one of the most challenging properties to infer.
% Method
Competing dating methods with different observational inputs often produce
    wildly inconsistent age estimates.
In this work we attempt to combine multiple methods into one coherent model
    and calibrate that model using galactic dynamics --- a well understood
    physical process.
We identify a sample of stars with at least two of the following dating
    methods available: rotation-dating, isochronal-dating,
    asteroseismic-dating and dynamical-dating.
% Results
\end{abstract}

@@

The processes behind the formation of the Milky Way and the objects within it
are some of the most elusive and complicated topics in astronomy today,
connected by a common theme: stellar ages.
Ages provide the key to understanding the evolution of all astrophysical
objects, but age is, unfortunately, one of the most difficult stellar
properties to measure.
Different dating methods often produce inconsistent predictions for the age of
the same star.
This is in part because the underlying processes generating the evolution of
the observable properties are different and in part because our understanding
of the underlying physics is flawed or incomplete.
The various available dating methods can be categorised by the underlying
physical process they trace --- the radial extent of the hydrogen-burning core
or the evolving magnetic dynamo for example.
In addition, they can be classified by their level of empiricism, \ie\ the
number of free parameters that need to tuned when calibrating these models.
The physics behind orbital heating, for example, is very well understood and
has only one tunable parameter.
On the other hand, magnetic activity evolution is poorly understood and the
dating methods that are based on it are often entirely empirical.
In table \ref{tab:dating_methods} we provide an overview of various dating
methods, the main observables associated with them, the underlying physics
driving the changing observables, the types of star the method applies to and
the empirical or physical nature of the model.

@@

In this paper we focus on four of these dating methods: rotation,
asteroseismology, isochrones and galactic dynamics.
A brief overview of each of these methods is provided in below.

@@

Main sequence (MS) stars comprise the majority of our galaxy but their ages
are notoriously difficult to measure.
Their positions on the HR diagram don't change significantly during their
    hydrogen burning lifetimes, a fact that is convenient for life on Earth
    but inconvenient for galactic archaeologists.
Now, due to the abundance of rotation periods for MS stars provided by Kepler
    and to-be provided by TESS, LSST and Wfirst, rotation-dating is the most
    readily available, precise method for inferring stellar ages.
Rotation-dating works well for young stars but a question mark still hangs
    over its accuracy for stars older than the Sun.
Recent results show that old \kepler\ asteroseismic stars rotate more rapidly
    than expected given their age \citep[\eg][]{Angus2015, Vansaders2016,
    Metcalfe2016}.
This has been attributed to an evolving magnetic dynamo: as stars reach an
    critical Rossby number (the ratio of rotation period to the convective
    overturn timescale), their magnetic field `switches off' and stars
    maintain a consistent rotation period after that time.
Whilst this physical explanation produces a model that fits the data, it
    is driven by observations, not theory, and other explanations could
    provide an answer.
The data sets typically used to test the age-rotation relations are highly
    heterogeneous and each set has its own detection and selection biases.
For example, asteroseismology favours quiet stars whereas rotation periods are
    easiest to measure for active stars.

@@

\subsection{Isochrones}

@@

>>>>

\begin{abstract}
% Aims
Stellar age underpins a wide range of astronomical applications but
remains one of the most challenging stellar properties to infer.
% Method
Competing dating methods with different observational inputs often produce
    wildly inconsistent age estimates.
In this work we attempt to combine multiple methods into one coherent model.
For the first time, we combine apparent magnitudes and colour information,
    typically used to infer an age using stellar evolutionary tracks, with
    other age proxies: rotation period and galactic kinematics.
We identify a sample of stars with rotation periods, colours, spectroscopic
    parameters and kinematics to calibrate this dating method.
% Results
\end{abstract}

@@

The processes behind the formation of the Milky Way and the objects within it
are some of the most elusive and complicated topics in astronomy today.
These topics are connected by a common theme: stellar ages.
Stellar ages provide the key to understanding the evolution of all
astrophysical objects.
For main sequence (MS) stars however, age is a difficult property to infer.
This is predominantly because hydrogen burning stars do not change appreciably
during their time spent on the MS: a star like the Sun will grow in
luminosity by around a factor of two before becoming a red giant.
Luminosity is therefore not a very sensitive probe of age.
As explained further later in this paper: the Sun's rotation period will vary
by almost an order of magnitude over its MS lifetime, thus rotation
period is a much more sensitive age proxy than luminosity or temperature.

In addition to the difficulties imposed by the slow timescale for variability
withing MS stars, different dating methods often produce inconsistent
predictions for the age of a star.
For example, an asteroseismic age will not necessarily agree with a isochronal
or rotational age.
This is in part because the underlying processes generating the evolution of
the observable properties are different and in part because our understanding
of the underlying physics is flawed or incomplete.
The various available dating methods can be categorised by the underlying
physical process they trace.
For example, evolutionary models track the radial extent of the
hydrogen-burning core and age-rotation relations model the evolving state of
the internal magnetic dynamo.
In addition, dating methods can be classified by their level of empiricism,
\ie\ the number of free parameters that need to tuned when fitting the models
to the data.
The physics behind the evolving luminosity and effective temperature of a star
as a result of core hydrogen burning is, for example, very well understood and
does not need calibrating; physics determine these models.
On the other hand, magnetic activity evolution is poorly understood and must
be calibrated using available data.
In table \ref{tab:dating_methods} we provide an overview of various dating
methods, the main observables associated with them, the underlying physics
driving the changing observables, the types of star the method applies to and
the empirical or physical nature of the model.

@@

In this paper we focus on four of these dating methods: rotation,
asteroseismology, stellar evolution models and galactic kinematics.
A brief overview of each of these methods is provided in the following
sections.

@@

Main sequence (MS) stars comprise the majority of our galaxy but their ages
are notoriously difficult to measure.
Their positions on the HR diagram don't change significantly during their
    hydrogen burning lifetimes, a fact that is convenient for life on Earth
    but inconvenient for galactic archaeologists.
Now, due to the abundance of rotation periods for MS stars provided by Kepler
    and to-be provided by TESS, LSST and Wfirst, rotation-dating is the most
    readily available, precise method for inferring stellar ages.
Rotation-dating works well for young stars but a question mark still hangs
    over its accuracy for stars older than the Sun.
Recent results show that old \kepler\ asteroseismic stars rotate more rapidly
    than expected given their age \citep[\eg][]{Angus2015, Vansaders2016,
    Metcalfe2016}.
This has been attributed to an evolving magnetic dynamo: as stars reach a
    critical Rossby number (the ratio of rotation period to the convective
    overturn timescale), their magnetic field `switches off' and stars
    maintain a consistent rotation period after that time.
Whilst this physical explanation produces a model that fits the data, it
    is driven by observations, not theory, and other explanations could
    provide an answer.
The data sets typically used to test the age-rotation relations are highly
    heterogeneous and each set has its own detection and selection biases.
For example, asteroseismology favours quiet stars whereas rotation periods are
    easiest to measure for active stars.

@@

\subsection{Stellar Evolution models}
The ongoing physical processes in the core of a star is reflected externally
by an increase in luminosity and temperature.
As hydrogen is converted to helium via nuclear fusion in the core, the mass
fraction, etc, etc.
Leading to etc, etc.

@@

We combine the following dating methods: rotation, stellar evolution models
and kinematics into one model via self calibration.
As an example, take the Sun: a star with temperature $5777K$, $ \log(g) =
4.44$ $[Fe/H] = 0.0$ and $P_{rot} = 26$ days.
